---
title: "Biogeography"
author: "Dean Adams (dcadams@iastate.edu)"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: ../../Instructor_Materials/style/styles.css
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r,echo=F}

knitr::read_chunk('Biogeography.R')

```

### **Overview**
Today we will explore how to perform a biogeography analysis in a phylogenetic framework. We learned about two models: DEC (dispersal-exctinction-cladogenesis) and DEC+J (dispersal-extinction-cladogenesis plus jumps). **NOTE: much of this tutorial follows the Supplemental tutorial from Chapter 11 in Revell and Harmon (Phylogenetic Comparative Methods in R).** 

To implement these models, we require the package `BioGeoBEARS`. This is not on CRAN, so we must install from Github: 

```{r, eval = F, echo=T}
devtools::install_github('nmatzke/BioGeoBEARS')
```


### Download data files
Today we will use data from the following files: 

[Phylogeny](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/data/attine-tree-pruned.tre)

[Data](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/data/attine-distribution-data.txt)

Additionally, The R-script for running the code found in this tutorial is found here:

[RScript](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/scripts/Biogeography.r)

As before, on your local computer, place them in a folder named 'TutorialData', which is found in the same directory as this tutorial. 

### **Reading Data**

First we will load libraries and read in the data. Today we will be examining data from Branstetter et al. (2017). These are data for fungus-farming ant species, whose biogeographic distributions are described as presence/absence in 5 geographic regions around the world. Our aim is to hypothesize the ancestral origin of this group.

```{r,read_data, eval=TRUE}

```

We see that our data consist of 84 species, found in one or more biogeographic regions (five regions total). 

### **Plotting the data**

As always, it is useful to plot the data. Here let's plot the biogeographic regions on the phylogeny: 

```{r,plot_data, eval=TRUE}

```

Notice that all species are geographically-restricted to one or two regions of the globe. 

### **DEC Analysis**

Now let's run a DEC analysis, where we describe biogeographic shifts across the phylogeny in terms of dispersal, extinction, and cladogenesis. One quirk with `BioGeoBEARS` is that unlike most `R`-packages where data are loaded into the `R` environment, This package reads directly from data files. So the input looks a bit different from what we've seen in earlier tutorials. Notice that on your computer, `BioGeoBEARS` spits out considerable output during the analysis. For ease of presentation that has been suppressed in this tutorial. 

```{r,run_dec, eval=TRUE, results='hide'}

```

```{r,res_dec, eval=TRUE}

```

Here is the optimized output: 

- p1 = the disperal rate
- p2 = local extinction rate
- value = logL

Now let's plot the results graphically: 

```{r,plot_dec, eval=TRUE}

```

This shows us that Region C (South America) is the most likely ancestral area for the group.

### **DEC+J Analysis**

Now let's consider the DEC+J model. This includes dispersal, extinction, cladogenesis, and jump-speciation (where a lineage disperses to a new location and simultaneously speciates). 

To implement the model, let's use our DEC parameters as the starting point for DEC+J: 

```{r,run_decj, eval=TRUE, results='hide'}

```

```{r,res_decj, eval=TRUE}

```

Here is the optimized output. As before, p1 = the disperal rate, p2 = local extinction rate, and now p3 = the jump speciation rate. 

Now let's plot the results graphically: 

```{r,plot_decj, eval=TRUE}

```

Notice that similar to the DEC model, the ancestral area for the group is hypothesized to be C (South America). However, hypothesized area reconstructions within the phylogeny do differ between the two models. 

### **Model Comparison**

Because some details differ we may wish to formally perform a model comparison to deterimine which model confers the stronger statistical support. 

```{r,model_compare, eval=TRUE}

```

Here we see that the DEC+J model has much stronger support than the DEC model. Thus, we reject DEC in favor of a model that also includes jump speciation. 