---
title: "Biogeography"
author: "Dean Adams (dcadams@iastate.edu)"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: ../../Instructor_Materials/style/styles.css
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---

```{r,echo=F}

knitr::read_chunk('Biogeography.R')

```

### **Overview**
Today we will explore how to perform a biogeography analysis in a phylogenetic framework. We learned about two models: DEC (dispersal-exctinction-cladogenesis) and DEC+J (dispersal-extinction-cladogenesis plus jumps). **NOTE: much of this tutorial follows the Supplemental tutorial from Chapter 11 in Revell and Harmon (Phylogenetic Comparative Methods in R).** 

To implement these models, we require the package `BioGeoBEARS`. This is not on CRAN, so we must install from Github: 

```{r, eval = F, echo=T}
devtools::install_github('nmatzke/BioGeoBEARS')
```


### Download data files
Today we will use data from the following files: 

[Phylogeny](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/data/attine-tree-pruned.tre)

[Data](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/data/attine-distribution-data.txt)

Additionally, The R-script for running the code found in this tutorial is found here:

[RScript](https://github.com/deanadams/MacroevolutionPracticals/blob/main/14-Biogeography/scripts/Biogeography.r)

As before, on your local computer, place them in a folder named 'TutorialData', which is found in the same directory as this tutorial. 

### **Reading Data**

First we will load libraries and read in the data. Today we will be examining data from Branstetter et al. (2017). These are data for fungus-farming ant species, whose biogeographic distributions are described as presence/absence in 5 geographic regions around the world. Our aim is to hypothesize the ancestral origin of this group.

```{r,read_data, eval=TRUE}

```

We see that our data consist of 84 species, found in one or more biogeographic regions (five regions total). 

### **Plotting the data**

As always, it is useful to plot the data. Here let's plot the biogeographic regions on the phylogeny: 

```{r,plot_data, eval=TRUE}

```

Notice that all species are geographically-restricted to one or two regions of the globe. 

### **DEC Analysis**

Now let's run a DEC analysis, where we describe biogeographic shifts across the phylogeny in terms of dispersal, extinction, and cladogenesis. 

